A Plasma Torus Around a Young Low-Mass Star

### 2025-01-18 paper/ms.tex
update

% ** keck/hires: emission beyond v/veq>1 (2-3 paragraphs)
The spectroscopy shows emission well outside the star's
equatorial velocity ($v_{\rm eq}$=130\,\kms).  There are at least two
distinct emission components, each spaced half a cycle apart in phase.  The
first has clearer sinusoidal behaviour and is double-peaked, with
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.  The
flux excesses from these two peaks are correlated with one another,
with amplitudes varying from 100\% of the continuum flux early in the
observation sequence to 30\% by its end.  The component 180$^\circ$
opposite in phase is only detected from $\phi$=0.2-1.0, and 
from $\phi$=0.2-0.5, this latter component appears connected to the
star in velocity space.  While its peak semi-amplitude of
$K_1$=3.9\,$v_{\rm eq}$ is achieved at both $\phi$=0.25 and 0.75, its
amplitude decreases from a 60\% excess beyond the continuum to a 10\%
excess.   The sinusoidal period in all cases where emission is seen is
consistent with the photometric \periodhr\ hour period.  As we shall
discuss below, the only plausible explanation for the behavior at
$|\Delta v / v_{\rm eq}|>1$ is that circumstellar clumps of hydrogen
are corotating with the star.  These clumps transit in front of the
star when passing from negative to positive velocity.

% * keck/hires: absorption within v/veq<1
% * keck/hires: emission within v/veq<1
Within the stellar H$\alpha$ line core, at $|\Delta v / v_{\rm
eq}|<1$, the behavior is more complex.  Generally, one would
expect this region to be generated by the net superposition of bright
and dark regions on the stellar surface, and then modulated by any
occulting material capable of absorbing or emitting in H$\alpha$.  In
Figure~\ref{fig:spec}e, the behavior from $\phi$=0.4-1.2, is most easily
interpreted: from $\phi$=0.4-0.9, a hot region first gradually crosses
the stellar line profile, following from $\phi$=0.7-1.2 by the transit
of a cool region.  The phases $\phi$$<$0.4 seem to be a mix of similar
events, though the time sampling is sufficiently coarse that the
interpretation is less clear.
A final exercise to quantify the behavior in the line core is
shown in Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes
the summed flux in the at $|\Delta v / v_{\rm eq}|<1$.
Changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
higher-velocity clump and the occultation of the lower-velocity clump.

@@

The sinusoidal emission features require clumps of partially-ionized
hydrogen to be corotating with the star.
The velocity semi-amplitude of the sinusoids directly gives the
distance of these clumps from the stellar surface: 2.1-2.7\,$R_\star$
for one clump, and 3.9\,$R_\star$ for the other.
Their motion, rather than being Keplerian, can only be explained by
plasma being dragged along with the rotating stellar magnetic field.

The new data rule out two plausible origin scenarios for CPVs.
The ``starspot-only'' scenario \cite{Koen2021} has no means of
explaining this type of spectroscopic emission.
Similarly, a ``dust-only'' scenario in which the circumstellar
material would have a pure dust composition is also ruled out:
to explain the H$\alpha$ emission, the circumstellar clumps need to be
made of either a pure plasma, or a dusty plasma.
A priori, there are three plausible sources of such material:
the star, an old and undetected disk, or outgassing rocky bodies.

% TODO TODO: the rest!
* role of B: comparison to, and need for, RMHD models
  * massive B star / jupiter cnxn

* more wavelengths:
    Radio detection.   Relevance for B measurement.
    UV?  xray?
    need for jwst

* microphysics: the (probable) need for dust
    W1,W2,W3 no dust evidence.
    W4 detection questionable based on WISE images.

@@

\subsection{Modeling the Emitting Clump}

>>>>

% ** keck/hires: emission beyond v/veq>1 (2-3 paragraphs)
The spectroscopy shows emission well outside the star's
equatorial velocity ($v_{\rm eq}$=130\,\kms).  There are at least two
distinct emission components, each spaced half a cycle apart in phase.  The
first has clearer sinusoidal behaviour and is double-peaked, with
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.  The
flux excesses from these two peaks are correlated with one another,
with amplitudes varying from 100\% of the continuum flux early in the
observation sequence to 30\% by its end.  The component 180$^\circ$
opposite in phase is only detected from $\phi$=0.2-1.0, and 
from $\phi$=0.2-0.5, this latter component appears connected to the
star in velocity space.  While its peak semi-amplitude of
$K_1$=3.9\,$v_{\rm eq}$ is achieved at both $\phi$=0.25 and 0.75, its
amplitude decreases from a 60\% excess beyond the continuum to a 10\%
excess.   The sinusoidal period in all cases where emission is seen is
consistent with the photometric \periodhr\ hour period.

The sinusoidal emission features require circumstellar clumps of
partially-ionized hydrogen to be corotating with the star.  The
velocity semi-amplitude of the sinusoids directly gives the distance
of these clumps from the stellar surface: 2.1-2.7\,$R_\star$ for the
closer clump, and 3.9\,$R_\star$ for the other.  This material's
motion, rather than being Keplerian, can only be explained by plasma
being dragged along with the rotating stellar magnetic field.  These
clumps transit in front of the star when passing from negative to
positive velocity.

% * keck/hires: absorption within v/veq<1
% * keck/hires: emission within v/veq<1
The behavior within the stellar H$\alpha$ line core, at $|\Delta v /
v_{\rm eq}|<1$, is more complex than outside it.  Generally, one would
expect the emission in the line core to be generated near the stellar
chromosphere, near the star's surface, and then modulated by any
occulting material capable of absorbing or emitting in H$\alpha$.  In
Figure~\ref{fig:spec}e, the behavior from $\phi$=0.4-1.2 is most
easy to interpret: from $\phi$=0.4-0.9, a hot region first gradually
crosses the stellar line profile, followed from $\phi$=0.7-1.2 by the
transit of a cool region.  The phases $\phi$$<$0.4 seem to be a mix of
similar events, though the time sampling is sufficiently coarse that
the interpretation is less clear.  A final exercise to quantify the
behavior in the line core is shown in Figure~\ref{fig:spec}d, where
$f_{\rm H\alpha\ core}$ denotes the summed flux in the at $|\Delta v /
v_{\rm eq}|<1$.  Changes in the line core flux are usually correlated
with the broadband variability, except at $\phi$=0.5, during the
transit of the higher-velocity clump and the occultation of the
lower-velocity clump.

@@

Magetically-active, rapidly rotating stars with a wide range of masses
have been known to exhibit both sinusoidal emission features (CITE,
CITE, CITE, CITE) as well as sharp transient absorption features in
their line cores (CITE, CITE, CITE, CITE) similar to those in
Figure~\ref{fig:spec}.
None of the low mass stars showing such spectroscopic features have
previously shown abnormal light curves (CITE).
The preferred interpretation for such spectroscopic variability comes
from a loose analogy to quiescent solar prominences and filaments,
which are cool condensations of plasma in the solar corona that last
days to weeks (CITE).
This plasma is called a prominence when viewed in emission against the
dark backdrop of space, and a filament when viewed in absorption
against the solar disk.
In our Sun's magnetosphere, these condensations fall back to
the solar surface because gravity is stronger than any magnetic or
centrifugal force capable of sustaining them.
However for stars with magnetospheric radii $R_{\rm m}$ that exceed the
corotation radii $R_{\rm c}$, the effective potential experienced by
the plasma have local minima outside the $R_{\rm c}$, enabling the
material to be sustained for much longer timescales (see CITE, CITE).

Our Keck/HIRES observations were the first time a CPV was observed
with sufficient sensitivity to test for the presence of circumstellar
plasma clumps, and they showed for TIC~141146667 that they exist and
are corotating with the star.  
Characteristic densities and masses of these clumps are $n \sim
10^{10}$\,cm$^{-3}$ and $M \sim 10^{14}$\,kg (see~\ref{subsec:model}),
a similar density to solar prominences, but ten to one hundred times
more massive.
This observation rules out a ``starspot-only'' origin scenario
for CPVs, \cite{Koen2021} since such scenarios have no means of
explaining spectroscopic emission beyond the stellar disk.  Similarly,
scenarios in which the circumstellar material is made only of dust are
also ruled out.  Although dust may be present, to explain the
H$\alpha$ emission, the circumstellar clumps must include plasma with
a significant population of hydrogen atoms in the $n$=3 excited state.
While the plasma is undoubtedly sculpted by the star's magnetic field,
there are three a priori plausible origin locations for the material:
the star, an old and undetected disk, or outgassing rocky bodies.

The most intriguing potential known analogs to CPVs 
are the $\sigma$~Ori~E variables, a rare subset of B stars that almost
all have tilted-dipole magnetic fields (CITE) which can trap outflowing
stellar winds into warped plasma tori (CITE). 
These tori tend to have dense antipodal accumulations of plasma, and
the transits of these accumulations can produce broadband photometric
variability through a combination of bound-free scattering (CITE) and
Thomson scattering (CITE).  For $\sigma$~Ori~E
and most of its analogs, the result is light curves that almost all
appear ``simple'', resembling those of eclipsing binaries (CITE,
CITE).  The two known exceptions, HD~37776 and HD~64740, show complex
light curves resembling CPVs (Mikulasek2020, CITEBouma2024), and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments (CITE CITE).  There are two
implications: firstly, the complexity of CPVs may be a direct
consequence of magnetic fields with highly multipolar contributions.
Secondly, CPVs could be a source of astrophysical false positives in
photometric searches for eclipsing binaries and transiting exoplanets
around young pre-main-sequence M dwarfs (e.g.  CITE).

% Need paragraph(s) on the future!
Pressing issues for future work on CPVs include determining the
composition and origin of the circumstellar material, understanding
the exact role of the stellar magnetic field, and exploring the
implied space weather experienced by the close-in rocky exoplanets
that, statistically, are likely to be nearby (CITE).

The composition -- either purely plasma, or else a dusty plasma -- can
be clarified by time-series optical and infrared spectrophotometry.
While observations of CPVs in the optical have previously suggested
the presence of dust (CITE), plasma clumps transiting over starspots
can also produce chromatic features (CITERackham).  The composition
and size distribution of any dust that is present could be most easily
resolved by measuring the extinction curve for one or more CPVs from
$\approx$1-10\,$\mu$m.  A composition similar to debris from rocky
bodies seen around white dwarfs (CITE Reach+09) would be indicative of
a rocky-body origin.  A composition closer to the ISM would be
indicative of condensed dust in an M dwarf wind, similar to that
formed in the environments of more evolved stars (CITE Marigo2008).

The role of the star's magnetic field could be better understood
through new observations, and new theory.
From the theoretical perspective, there is an urgent need for
rigid-field magnetohydrodynamic (MHD) modeling to go beyond previous work
(CITE CITE) by exploring the effects of non-dipolar field
contributions.
Fully time-dynamic MHD (e.g. CITE DaleyYates) will offer the
capability of understanding the connection between the plasma and the
dust.  %TODO: need this sentence to be better
Observationally, optical spectropolarimetry on sufficiently large
telescopes has the potential to assess both the field strength and
topology.
A more direct probe however might be to search for radio emission from
mechanisms such as the electron cyclotron maser instability, which can
provide a direct measurement of the field strength at the site of the
emitting region (typically an auroral oval).
Recent work (CITE) has shown that CPVs are variable radio emitters,
and that they show emission components that can be both persistent, as
well as short-lived and highly polarized.
%Observation in the x-ray would similarly inform our view of the hotter
%plasma in these magnetospheres which otherwise remains undetected.

It is currently unclear what, if any, relationship CPVs have to the
close-in exoplanets that exist around most M dwarfs (CITE).  However,
0.3-3\% of young M dwarfs show the CPV phenomenon (CITE),
and the circumstellar clumps of material need to be viewed
edge-on for the light curves to be ``complex''.  This implies that an
appreciable minority (3-30\%) of M dwarfs -- the rapidly rotating ones
with centrifugal magnetospheres -- seem to have similar circumstellar
environments to the CPVs.

@@

\subsection{Modeling the Emitting Clump}
\label{subsec:model}
